---
title: 'isdbayes: Bayesian hierarchical modeling of size spectra'
author: "Jeff Wesner"
output: 
  github_document
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This package allows the estimation of power law exponents using the truncated (upper and lower) Pareto distribution [@wesner2023bayesian]. Specifically, it allows users to fit Bayesian (non)-linear hierarchical models with a truncated Pareto likelihood using `brms` [@burkner2017brms]. The motivation for the package was to estimate power law exponents of ecological size spectra using individual-level body size data in a generalized mixed model framework. The likelihood for the truncated Pareto used here was described in [@edwards2020]. This package translates that likelihood into `brms`.

## Installation

This package requires installation of `brms` and `rstan`, which itself requires installation of a C++ toolchain.

1)  Go to <https://mc-stan.org/users/interfaces/rstan.html> and follow the instructions to install `rstan` and configure the C++ toolchain.

2)  Install the latest version of `brms` with install.packages("brms").

3)  Install `isdbayes` from github using `devtools`:

```{r, eval = F}
# requires an installation of devtools

devtools::install_github("jswesner/isdbayes")

```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
devtools::load_all()
library(tidyverse)
library(tidybayes)
```


# Examples
## Fit individual samples
First, simulate some power law data using `rparetocounts()`. The code below simulates 300 body sizes from a power law with exponent (mu) = -1.2, xmin (vreal2) = 1, and xmax (vreal3) = 1000. The options are called "mu", "vreal2", and "vreal3" instead of "lambda", "xmin", and "xmax" to fit with the generic naming requirements of `brms` for custom family distributions.

```{r, eval = F}
library(isdbayes)

# simulate data
dat = tibble(x = rparetocounts(n = 300,  mu = -1.2,  vreal2 = 1, vreal3 = 1000)) %>% 
  mutate(xmin = min(x),
         xmax = max(x),
         counts = 1)

```

The code above simulates data from a doubly-truncated Pareto and then estimates xmin and xmax. It also adds a column for *counts.* If the data all represent unique individual masses, then this column takes a value of 1 for every body size. If the data have repeated sizes, then this column can take an integer or double of the counts or densities of those sizes. For example, data that are x = {1.9, 1.9, 1.8, 2.8, 2.8} could either be analyzed with each body size assumed to be unique where counts = {1, 1, 1, 1, 1} or it could be analyzed as x = {1.9, 1.8, 2.8} and counts = {2, 1, 2}. The latter is a common format when there is a density estimate associated with counts or a sampling effort.

Next estimate the power law exponent using `brms`.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
fit1 = readRDS(here::here("models/fit1.rds"))
fit2 = readRDS(here::here("models/fit2.rds"))
fit3 = readRDS(here::here("models/fit3.rds"))
```


```{r, eval = F}
library(brms)

fit1 = brm(x | vreal(counts, xmin, xmax) ~ 1, 
          data = dat,
          stanvars = stanvars,    # required for truncated Pareto
          family = paretocounts(),# required for truncated Pareto
          chains = 1, iter = 1000)

```

This example fits an intercept-only model to estimate the power-law exponent. For more complex examples with fixed and hierarchical predictors, see below.


## Simulate multiple size distributions
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(isdbayes)
library(tidyverse)
library(brms)

x1 = rparetocounts(mu = -1.8) # `mu` is required wording from brms. in this case it means the lambda exponent of the ISD
x2 = rparetocounts(mu = -1.5)
x3 = rparetocounts(mu = -1.2)

isd_data = tibble(x1 = x1,
                  x2 = x2,
                  x3 = x3) %>% 
  pivot_longer(cols = everything(), names_to = "group", values_to = "x") %>% 
  group_by(group) %>% 
  mutate(xmin = min(x),
         xmax = max(x)) %>% 
  group_by(group, x) %>% 
  add_count(name = "counts")
```

## Fit multiple size distributions with a fixed factor
```{r, eval = F}
fit2 = brm(x | vreal(counts, xmin, xmax) ~ group, 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 1, iter = 1000)
```

## Plot group posteriors
```{r}
posts_group = fit2$data %>% 
  distinct(group, xmin, xmax) %>% 
  mutate(counts = 1) %>% 
  tidybayes::add_epred_draws(fit2, re_formula = NA) 

posts_group %>% 
  ggplot(aes(x = group, y = .epred)) + 
  tidybayes::stat_halfeye(scale = 0.2) + 
  geom_hline(yintercept = c(-1.8, -1.5, -1.2)) # known lambdas
```

## Fit multiple size distributions with a varying intercept
```{r, eval = F}
fit3 = brm(x | vreal(counts, xmin, xmax) ~ (1|group), 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 1, iter = 1000)
```
## Plot varying intercepts
```{r}
posts_varint = fit3$data %>% 
  distinct(group, xmin, xmax) %>% 
  mutate(counts = 1) %>% 
  tidybayes::add_epred_draws(fit3, re_formula = NULL) 

posts_varint %>% 
  ggplot(aes(x = group, y = .epred)) + 
  tidybayes::stat_halfeye(scale = 0.2) + 
  geom_hline(yintercept = c(-1.8, -1.5, -1.2)) # known lambdas
```


## Posterior predictive checks
After the model is fit, you can use built-in functions in brms to perform model checking.
```{r}
pp_check(fit2, type = "dens_overlay_grouped", group = "group")
```

## References

::: {#refs}
:::
