---
title: 'isdbayes: Bayesian hierarchical modeling of size spectra'
author: "Jeff Wesner"
date: "2023-10-04"
output: github_document
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This package allows the estimation of power law exponents using the truncated (upper and lower) Pareto distribution [@wesner2023bayesian]. Specifically, it allows users to fit Bayesian (non)-linear hierarchical models with a truncated Pareto likelihood using `brms` [@burkner2017brms]. The motivation for the package was to estimate power law exponents of ecological size spectra using individual-level body size data in a generalized mixed model framework. The likelihood for the truncated Pareto used here was described in [@edwards2020]. This package translates that likelihood into `brms`.

## Installation

This package requires installation of `brms` and `rstan`, which itself requires installation of a C++ toolchain.

1)  Go to <https://mc-stan.org/users/interfaces/rstan.html> and follow the instructions to install `rstan` and configure the C++ toolchain.

2)  Install the latest version of `brms` with install.packages("brms").

3)  Install `isdbayes` from github using `devtools`:

```{r, eval = F}
# requires an installation of devtools

devtools::install_github("jswesner/isdbayes")

```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
devtools::load_all()
library(tidyverse)
library(tidybayes)
```


# Examples
## Fit individual samples
`isdbayes` amends the stanvars and family options in brms models to accept the truncated Pareto. First, simulate some power law data using the `rparetocounts()` function. The code below simulates 300 body sizes from a power law with exponent (mu) = -1.2, xmin (vreal2) = 1, and xmax (vreal3) = 1000. The options are called "mu", "vreal2", and "vreal3" instead of "lambda", "xmin", and "xmax" to fit with the generic naming requirements of `brms` for custom family distributions.

```{r, eval = F}
library(isdbayes)

# simulate data
dat = tibble(x = rparetocounts(n = 300,  mu = -1.2,  vreal2 = 1, vreal3 = 1000)) %>% 
  mutate(xmin = min(x),
         xmax = max(x),
         counts = 1)

```

The code above simulates data from a doubly-truncated Pareto and then estimates xmin and xmax. It also adds a column for *counts.* If the data all represent unique individual masses, then this column takes a value of 1 for every body size. If the data have repeated sizes, then this column can take an integer of the counts of those sizes. For example, data that are x = {1.9, 1.9, 1.8, 2.8, 2.8} could either be analyzed with each body size assumed to be unique where counts = {1, 1, 1, 1, 1} or it could be analyzed as x = {1.9, 1.8, 2.8} and counts = {2, 1, 2}. The latter is a common format when there is a density estimate associated with counts or a sampling effort.

Next estimate the power law exponent using `brms`.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
fit1 = readRDS(here::here("models/fit1.rds"))
fit2 = readRDS(here::here("models/fit2.rds"))
fit3 = readRDS(here::here("models/fit3.rds"))
```


```{r, eval = F}
library(brms)

fit1 = brm(x | vreal(counts, xmin, xmax) ~ 1, 
          data = dat,
          stanvars = stanvars,    # required for truncated Pareto
          family = paretocounts(),# required for truncated Pareto
          chains = 1, iter = 1000)

```

This example fits an intercept-only model to estimate the power-law exponent. For more complex examples with fixed and hierarchical predictors, see below.


# Simulate multiple size distributions
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(isdbayes)
library(tidyverse)
library(brms)

x1 = rparetocounts(mu = -1.8) # `mu` is required wording from brms. in this case it means the lambda exponent of the ISD
x2 = rparetocounts(mu = -1.5)
x3 = rparetocounts(mu = -1.2)

isd_data = tibble(x1 = x1,
                  x2 = x2,
                  x3 = x3) %>% 
  pivot_longer(cols = everything(), names_to = "group", values_to = "x") %>% 
  group_by(group) %>% 
  mutate(xmin = min(x),
         xmax = max(x)) %>% 
  group_by(group, x) %>% 
  add_count(name = "counts")
```

# Fit multiple size distributions with a fixed factor
```{r, eval = F}
fit2 = brm(x | vreal(counts, xmin, xmax) ~ group, 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 1, iter = 1000)
```

# Plot group posteriors
```{r}
posts_group = fit2$data %>% 
  distinct(group, xmin, xmax) %>% 
  mutate(counts = 1) %>% 
  tidybayes::add_epred_draws(fit2, re_formula = NA) 

posts_group %>% 
  ggplot(aes(x = group, y = .epred)) + 
  tidybayes::stat_halfeye(scale = 0.2) + 
  geom_hline(yintercept = c(-1.8, -1.5, -1.2)) # known lambdas
```

# Fit multiple size distributions with a varying intercept
```{r, eval = F}
fit3 = brm(x | vreal(counts, xmin, xmax) ~ (1|group), 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 1, iter = 1000)
```
# Plot varying intercepts
```{r}
posts_varint = fit3$data %>% 
  distinct(group, xmin, xmax) %>% 
  mutate(counts = 1) %>% 
  tidybayes::add_epred_draws(fit3, re_formula = NULL) 

posts_varint %>% 
  ggplot(aes(x = group, y = .epred)) + 
  tidybayes::stat_halfeye(scale = 0.2) + 
  geom_hline(yintercept = c(-1.8, -1.5, -1.2)) # known lambdas
```


# Posterior predictive checks
After the model is fit, you can use built-in functions in brms to perform model checking.
```{r}
pp_check(fit2, type = "dens_overlay_grouped", group = "group")
```

# Plotting the ISD
To plot the posterior against data, use plot_isd_posts() and then add data with get_isd_data(). These plots follow the format of @edwards2017testing, with body size values on the x-axis and the number of values $\geq$ x ("n_yx") on the y-axis. 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# wrangle isd data for plotting
isd_dat = make_isd_data(fit2, group = group)

# plot posterior ISD with data overlaid
plot_isd_posts(fit2, group = group) + 
  geom_point(data = isd_dat)
```

In the plot above, the posteriors are summarized with a median and 95% Credible Interval. To access the underling posteriors, use get_isd_posts()
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# get ISD posteriors
isd_posts = get_isd_posts(fit2, group = group)

head(isd_posts)
```

Now we can plot the posterior draws individually.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# get ISD posteriors
isd_posts %>% 
  ggplot(aes(x = x, y = n_yx, group = interaction(group, .draw), color = group)) +
  geom_line()
```

Adding log-scales to x
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# get ISD posteriors
isd_posts %>% 
  ggplot(aes(x = x, y = n_yx, group = interaction(group, .draw), color = group)) +
  geom_line() +
  scale_x_log10()
```

Adding log-scales to x and y requires cutting off at n_yx = 1 to avoid plotting errors
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# get ISD posteriors
isd_posts %>% 
  ggplot(aes(x = x, y = n_yx, group = interaction(group, .draw), color = group)) +
  geom_line() +
  scale_x_log10() +
  scale_y_log10() + 
  coord_cartesian(ylim = c(1, NA)) # this cuts of the y-axis. For the full posterior plot, remove scale_y_log10() the last line
```
## References

::: {#refs}
:::
