---
title: "isdbayes_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{isdbayes_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(isdbayes)
library(here)
library(brms)
library(tidyverse)
```

# simulate a single size distribution
This simulates data with rparetocounts(), which takes values of n, mu, vreal2, and vreal3. n is the number of body sizes to simulate, mu is lambda, vreal2 is xmin, and vreal3 is xmax. The use of "mu" and "vreal2" and "vreal3" is required by brms to represent the quantities of interest, so those terms are used in this package (which can be confusing).

```{r}
isd_data = tibble(x = rparetocounts()) %>% 
  mutate(xmin = min(x), # xmin
         xmax = max(x)) %>% #xmax
  group_by(x) %>% 
  add_tally(name = "counts") # get counts of each body size. If all sizes are unique, this is just 1. This is where size densities would also go
```

# estimate lambda
```{r}
fit1 = brm(x | vreal(counts, xmin, xmax) ~ 1, 
          data = isd_data,
          stanvars = stanvars,
          family = paretocounts(),
          chains = 1, iter = 1000)
```

# simulate multiple size distributions
```{r}

x1 = rparetocounts(mu = -1.8) # `mu` is required wording from brms. in this case it means the lambda exponent of the ISD
x2 = rparetocounts(mu = -1.5)
x3 = rparetocounts(mu = -1.2)

isd_data = tibble(x1 = x1,
                  x2 = x2,
                  x3 = x3) %>% 
  pivot_longer(cols = everything(), names_to = "group", values_to = "x") %>% 
  group_by(group) %>% 
  mutate(xmin = min(x),
         xmax = max(x)) %>% 
  group_by(group, x) %>% 
  add_count(name = "counts")
```

# fit multiple size distributions with a fixed factor
```{r}
fit2 = brm(x | vreal(counts, xmin, xmax) ~ group, 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 1, iter = 1000)
```

# plot group posteriors
```{r}
posts_group = fit2$data %>% 
  distinct(group, xmin, xmax) %>% 
  mutate(counts = 1) %>% 
  tidybayes::add_epred_draws(fit2, re_formula = NA) 

posts_group %>% 
  ggplot(aes(x = group, y = .epred)) + 
  tidybayes::stat_halfeye(scale = 0.2) + 
  geom_hline(yintercept = c(-1.8, -1.5, -1.2)) # known lambdas
```

# fit multiple size distributions with a varying intercept
```{r}
fit3 = brm(x | vreal(counts, xmin, xmax) ~ (1|group), 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 1, iter = 1000)
```
# plot varying intercepts
```{r}
posts_varint = fit3$data %>% 
  distinct(group, xmin, xmax) %>% 
  mutate(counts = 1) %>% 
  tidybayes::add_epred_draws(fit3, re_formula = NULL) 

posts_varint %>% 
  ggplot(aes(x = group, y = .epred)) + 
  tidybayes::stat_halfeye(scale = 0.2) + 
  geom_hline(yintercept = c(-1.8, -1.5, -1.2)) # known lambdas
```


# Posterior predictive checks
After the model is fit, you can use built-in functions in brms to perform model checking.
```{r}
pp_check(fit2, type = "dens_overlay_grouped", group = "group")
```

# Model Comparison
(This needs to be verified) You can also perform model selection:
```{r}
WAIC(fit2)
WAIC(fit3)

```


# plot the isd
Lambda represents the rate of decline in relative abundance as body size increases, but because it doesn't have units like hour, height, etc., it is somewhat abstract. It can be easier to see the variation by plotting the probability that a body size X is greater than or equal to another body size x. This code produces the posterior distribution of these so-called "isd plots."

```{r}
# get posterior predictions using add_epred_draws
fit2_posts = fit2$data %>% 
  distinct(group) %>% # optional groups
  mutate(counts = 1, # placeholder
         xmin = 1, # placeholder
         xmax = 1000) %>% # placeholder
  tidybayes::add_epred_draws(fit2)

# get posterior of prob X>=x
prob_yx_posts = get_isd_posts(fit2_posts)


# plot
prob_yx_posts %>% 
  ggplot(aes(x = x.PLB, y = prob_yx)) +
  geom_line(aes(group = interaction(group, .draw))) +
  facet_grid(~group) +
  scale_x_log10() +
  scale_y_log10()

```

# add data to isd plot
```{r}
# get prob X>=x or Num X>=x from data
isd_data = fit2$data %>%
  group_by(group) %>% 
  get_isd_data(resp = x)


# plot
prob_yx_posts %>% 
  ggplot(aes(x = x.PLB, y = prob_yx)) +
  geom_line(aes(group = interaction(group, .draw))) +
  facet_grid(~group) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(data = isd_data, aes(x = x), color = "yellow", size = 1)
```

